---
title: "Direct H5AD Loading with LoadH5AD"
output:
  rmarkdown::html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Direct H5AD Loading with LoadH5AD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(reticulate)

# Configure Python - try multiple options for portability
python_configured <- FALSE

# Option 1: ov-nightly conda env (local, working squidpy on ARM Macs)
if (!python_configured) {
  tryCatch({
    ov_python <- "/opt/homebrew/Caskroom/miniconda/base/envs/ov-nightly/bin/python"
    if (file.exists(ov_python)) {
      use_python(ov_python, required = TRUE)
      python_configured <- TRUE
    }
  }, error = function(e) NULL)
}

# Option 2: conda base (local)
if (!python_configured) {
  tryCatch({
    if (file.exists("/opt/homebrew/Caskroom/miniconda/base/bin/python3")) {
      use_python("/opt/homebrew/Caskroom/miniconda/base/bin/python3", required = TRUE)
      python_configured <- TRUE
    }
  }, error = function(e) NULL)
}

# Option 3: system python (CI/GitHub Actions)
if (!python_configured) {
  tryCatch({
    use_python(Sys.which("python3"), required = FALSE)
  }, error = function(e) NULL)
}

# Check if scanpy is available for conditional evaluation
has_scanpy <- py_module_available("scanpy")

# Check if SeuratData is available
has_SeuratData <- requireNamespace("SeuratData", quietly = TRUE)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  dev = "png"
)
```

```{r cleanup, echo=FALSE, results='hide'}
outs <- c("pbmc3k_direct.h5Seurat", "pbmc3k_direct.h5ad")
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```

## Introduction

srtdisk provides two ways to load h5ad files into R as Seurat objects:

1. **`Convert()` + `LoadH5Seurat()`** — the traditional two-step approach via an intermediate h5Seurat file
2. **`LoadH5AD()`** — direct loading without intermediate files (new in v0.3.0)

`LoadH5AD()` is simpler and faster for common use cases: load a scanpy-processed h5ad file directly into a Seurat object, preserving expression data, metadata, dimensional reductions, and neighbor graphs.

```{r packages}
library(Seurat)
library(srtdisk)
```

## Loading a CellxGene h5ad file

The package bundles a colorectal cancer sample from [CellxGene](https://cellxgene.cziscience.com) (935 cells x 25,344 genes). Let's load it directly with `LoadH5AD()`:

```{r load_crc}
h5ad_path <- system.file("testdata", "crc_sample.h5ad", package = "srtdisk")
crc <- LoadH5AD(h5ad_path, verbose = TRUE)
crc
```

The h5ad file is loaded in a single call. Let's inspect the result:

```{r inspect_crc}
# Assays and layers
cat("Assays:", paste(Assays(crc), collapse = ", "), "\n")
cat("Layers:", paste(Layers(crc), collapse = ", "), "\n")

# Metadata columns
cat("Metadata columns:", ncol(crc[[]]), "\n")
head(crc[[]][, 1:min(5, ncol(crc[[]]))], 3)
```

```{r inspect_crc_reductions}
# Dimensional reductions
cat("Reductions:", paste(names(crc@reductions), collapse = ", "), "\n")

# Neighbor graphs
if (length(crc@graphs) > 0) {
  cat("Graphs:", paste(names(crc@graphs), collapse = ", "), "\n")
}
```

### Visualize

If the h5ad file contains UMAP coordinates (stored in `obsm/X_umap`), they are automatically restored:

```{r plot_crc_umap}
if ("umap" %in% names(crc@reductions)) {
  DimPlot(crc, reduction = "umap", group.by = "tissue", pt.size = 0.5)
}
```

## Comparison: LoadH5AD vs Convert + LoadH5Seurat

The traditional approach requires two steps and creates an intermediate h5Seurat file:

```{r traditional, eval = file.exists(h5ad_path)}
# Traditional approach (two steps)
local_h5ad <- file.path(tempdir(), basename(h5ad_path))
file.copy(h5ad_path, local_h5ad, overwrite = TRUE)
Convert(local_h5ad, dest = "h5seurat", overwrite = TRUE)
h5seurat_path <- sub("\\.h5ad$", ".h5seurat", local_h5ad)
crc_traditional <- LoadH5Seurat(h5seurat_path)

cat("LoadH5AD cells:", ncol(crc), "\n")
cat("Traditional cells:", ncol(crc_traditional), "\n")
cat("LoadH5AD features:", nrow(crc), "\n")
cat("Traditional features:", nrow(crc_traditional), "\n")
```

Both approaches produce equivalent Seurat objects. `LoadH5AD()` is simpler when you don't need the intermediate h5Seurat file.

```{r cleanup_traditional, echo=FALSE, results='hide'}
local_copy <- if (exists("local_h5ad")) local_h5ad else h5ad_path
f <- sub("\\.h5ad$", ".h5seurat", local_copy)
if (exists("local_h5ad") && file.exists(local_h5ad)) file.remove(local_h5ad)
if (file.exists(f)) file.remove(f)
```

## Loading a Seurat-originated h5ad

You can also use `LoadH5AD()` to round-trip data that was originally created in Seurat, exported to h5ad via `Convert()`, and optionally processed in Python.

```{r roundtrip, eval = has_SeuratData}
library(SeuratData)
if ("pbmc3k.final" %in% rownames(InstalledData())) {
  data("pbmc3k.final", package = "pbmc3k.SeuratData")
  pbmc <- UpdateSeuratObject(pbmc3k.final)

  # Export to h5ad
  SaveH5Seurat(pbmc, filename = "pbmc3k_direct.h5Seurat", overwrite = TRUE)
  Convert("pbmc3k_direct.h5Seurat", dest = "h5ad", overwrite = TRUE)

  # Load back with LoadH5AD
  pbmc_loaded <- LoadH5AD("pbmc3k_direct.h5ad")
  pbmc_loaded
} else {
  message("pbmc3k.final dataset not installed; skipping roundtrip example.")
}
```

```{r roundtrip_compare, eval = has_SeuratData}
if (exists("pbmc", inherits = FALSE) && exists("pbmc_loaded", inherits = FALSE)) {
  cat("Original cells:", ncol(pbmc), "\n")
  cat("Loaded cells:", ncol(pbmc_loaded), "\n")

  # Check reductions
  cat("Original reductions:", paste(names(pbmc@reductions), collapse = ", "), "\n")
  cat("Loaded reductions:", paste(names(pbmc_loaded@reductions), collapse = ", "), "\n")
}
```

```{r roundtrip_plot, eval = has_SeuratData}
if (exists("pbmc_loaded", inherits = FALSE) && "umap" %in% names(pbmc_loaded@reductions)) {
  DimPlot(pbmc_loaded, reduction = "umap", group.by = "seurat_annotations",
          label = TRUE, pt.size = 0.5) + NoLegend()
}
```

## What LoadH5AD preserves

`LoadH5AD()` reads the following from h5ad files:

| h5ad Location | Seurat Destination | Description |
|--------------|-------------------|-------------|
| `X` | Default assay `data` layer | Expression matrix (sparse or dense) |
| `raw/X` | `counts` layer | Raw counts if present |
| `layers/*` | Additional layers | Named layers mapped to Seurat slots |
| `obs` | `meta.data` | Cell metadata (categorical preserved as factors) |
| `var` | Feature metadata | Gene-level annotations |
| `var['highly_variable']` | `VariableFeatures()` | Variable feature selection |
| `obsm/X_umap` | `reductions$umap` | UMAP coordinates |
| `obsm/X_pca` | `reductions$pca` | PCA embeddings |
| `obsm/X_tsne` | `reductions$tsne` | tSNE coordinates |
| `obsm/spatial` | Spatial coordinates | Via `ConvertH5ADSpatialToSeurat()` |
| `obsp/connectivities` | `graphs$RNA_snn` | SNN graph |
| `obsp/distances` | `graphs$RNA_nn` | Distance graph |
| `uns/*` | `misc` | Unstructured annotations |

## When to use LoadH5AD vs Convert

| Scenario | Recommended |
|----------|------------|
| Quick exploration of an h5ad file | `LoadH5AD()` |
| Round-trip editing (load, modify, re-export) | `Convert()` + `LoadH5Seurat()` |
| Need h5Seurat for other tools | `Convert()` |
| Loading scanpy-processed data for Seurat analysis | `LoadH5AD()` |
| Working with spatial h5ad from CellxGene | `LoadH5AD()` |

## Session Info

```{r sessionInfo}
sessionInfo()
```

```{r cleanup-final, echo=FALSE, results='hide'}
outs <- c("pbmc3k_direct.h5Seurat", "pbmc3k_direct.h5ad")
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```
