---
title: "Multimodal Data with MuData (h5mu)"
output:
  rmarkdown::html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Multimodal Data with MuData (h5mu)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(reticulate)

# Configure Python - try multiple options for portability
python_configured <- FALSE

# Option 1: ov-nightly conda env
if (!python_configured) {
  tryCatch({
    ov_python <- "/opt/homebrew/Caskroom/miniconda/base/envs/ov-nightly/bin/python"
    if (file.exists(ov_python)) {
      use_python(ov_python, required = TRUE)
      python_configured <- TRUE
    }
  }, error = function(e) NULL)
}

# Option 2: conda base
if (!python_configured) {
  tryCatch({
    if (file.exists("/opt/homebrew/Caskroom/miniconda/base/bin/python3")) {
      use_python("/opt/homebrew/Caskroom/miniconda/base/bin/python3", required = TRUE)
      python_configured <- TRUE
    }
  }, error = function(e) NULL)
}

# Option 3: system python (CI/GitHub Actions)
if (!python_configured) {
  tryCatch({
    use_python(Sys.which("python3"), required = FALSE)
  }, error = function(e) NULL)
}

has_muon <- py_module_available("muon")
has_SeuratData <- requireNamespace("SeuratData", quietly = TRUE)
has_MuDataSeurat <- requireNamespace("MuDataSeurat", quietly = TRUE)
can_run <- has_SeuratData && has_MuDataSeurat

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  out.width = "100%",
  dev = "png"
)
```

```{r cleanup, echo=FALSE, results='hide'}
outs <- c("cbmc_multimodal.h5mu", "cbmc_roundtrip.h5mu")
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```

## Introduction

The [MuData](https://muon.readthedocs.io/en/latest/io/mudata.html) format (`.h5mu`) stores multimodal single-cell data — multiple modalities (e.g., RNA + protein) in a single file with shared cell annotations. This is the native format for [muon](https://muon-tutorials.readthedocs.io/) in Python and [MuDataSeurat](https://pmbio.github.io/MuDataSeurat/) in R.

srtdisk v0.3.0 adds `SaveH5MU()` and `LoadH5MU()` to simplify working with h5mu files in Seurat workflows:

- **`SaveH5MU()`**: Export a multi-assay Seurat object to h5mu, with automatic assay-to-modality name mapping
- **`LoadH5MU()`**: Load an h5mu file as a Seurat object, with automatic modality-to-assay name mapping

### When to use h5mu vs h5ad

| Format | Use case |
|--------|----------|
| **h5ad** | Single-modality data (RNA only) |
| **h5mu** | Multi-modal data (RNA + ADT, RNA + ATAC, etc.) |

```{r packages}
library(Seurat)
library(srtdisk)
```

## Create a multimodal Seurat object

We use the `cbmc` dataset from SeuratData — a CITE-seq experiment with RNA and ADT (antibody-derived tag) assays:

```{r load_cbmc, eval = can_run}
library(SeuratData)
if (!"cbmc" %in% rownames(InstalledData())) {
  InstallData("cbmc")
}
data("cbmc", package = "cbmc.SeuratData")
cbmc <- UpdateSeuratObject(cbmc)

# Remove ADT features that overlap with RNA (MuDataSeurat requires unique feature names)
overlap <- intersect(rownames(cbmc[["ADT"]]), rownames(cbmc[["RNA"]]))
if (length(overlap) > 0) {
  adt_keep <- setdiff(rownames(cbmc[["ADT"]]), overlap)
  cbmc[["ADT"]] <- subset(cbmc[["ADT"]], features = adt_keep)
  cat("Removed", length(overlap), "overlapping features from ADT\n")
}

cat("Assays:", paste(Assays(cbmc), collapse = ", "), "\n")
cat("Cells:", ncol(cbmc), "\n")
cat("RNA features:", nrow(cbmc[["RNA"]]), "\n")
cat("ADT features:", nrow(cbmc[["ADT"]]), "\n")
```

This object has two assays — RNA (gene expression) and ADT (surface protein) — measured on the same cells, making it ideal for h5mu.

> **Note**: MuDataSeurat requires unique feature names across modalities. If your CITE-seq ADT panel has features that share names with RNA genes (e.g., CD3D), remove duplicates before saving.

## Save with SaveH5MU

`SaveH5MU()` exports all assays to a single h5mu file. Assay names are automatically mapped to standard MuData modality names:

| Seurat Assay | h5mu Modality |
|-------------|---------------|
| RNA | rna |
| ADT | prot |
| ATAC | atac |
| Spatial | spatial |
| Other | lowercase name |

```{r save_h5mu, eval = can_run}
SaveH5MU(cbmc, "cbmc_multimodal.h5mu", overwrite = TRUE)
```

### Inspect in Python

If you have [muon](https://muon-tutorials.readthedocs.io/) installed, you can inspect the h5mu file:

```{python inspect_h5mu, eval = can_run && has_muon}
import muon as mu
mdata = mu.read("cbmc_multimodal.h5mu")
print(mdata)
print("\nModalities:", list(mdata.mod.keys()))
for mod_name, mod in mdata.mod.items():
    print(f"  {mod_name}: {mod.n_obs} cells x {mod.n_vars} features")
```

## Load with LoadH5MU

`LoadH5MU()` reads the h5mu file back into a Seurat object, reversing the modality name mapping:

| h5mu Modality | Seurat Assay |
|---------------|-------------|
| rna | RNA |
| prot | ADT |
| atac | ATAC |
| spatial | Spatial |
| Other | preserved |

```{r load_h5mu, eval = can_run}
cbmc_loaded <- LoadH5MU("cbmc_multimodal.h5mu")
cbmc_loaded
```

## Verify round-trip preservation

```{r verify_roundtrip, eval = can_run}
cat("=== Original ===\n")
cat("Assays:", paste(Assays(cbmc), collapse = ", "), "\n")
cat("Cells:", ncol(cbmc), "\n")

cat("\n=== Loaded ===\n")
cat("Assays:", paste(Assays(cbmc_loaded), collapse = ", "), "\n")
cat("Cells:", ncol(cbmc_loaded), "\n")

# Check metadata
cat("\nMetadata columns (original):", ncol(cbmc[[]]), "\n")
cat("Metadata columns (loaded):", ncol(cbmc_loaded[[]]), "\n")
```

## Custom modality name mapping

You can override the default mapping with `modality.names` (for `SaveH5MU()`) or `assay.names` (for `LoadH5MU()`):

```{r custom_mapping, eval = can_run}
# Save with custom modality names
SaveH5MU(
  cbmc,
  "cbmc_roundtrip.h5mu",
  modality.names = c(RNA = "gene_expression", ADT = "surface_protein"),
  overwrite = TRUE
)

# Load with custom assay names
cbmc_custom <- LoadH5MU(
  "cbmc_roundtrip.h5mu",
  assay.names = c(gene_expression = "RNA", surface_protein = "Protein")
)

cat("Assays:", paste(Assays(cbmc_custom), collapse = ", "), "\n")
```

## Conflict resolution

When a Seurat object has multiple assays that would map to the same modality name, `SaveH5MU()` automatically resolves the conflict. For example, if you have both RNA and SCT assays (both map to "rna" by default), SCT is remapped to "sct":

| Seurat Assays | Modality Mapping |
|--------------|-----------------|
| RNA, ADT | rna, prot (no conflict) |
| RNA, SCT, ADT | rna, sct, prot (SCT remapped) |

This ensures each modality has a unique name in the h5mu file.

## Data mapping reference

### SaveH5MU: Seurat to h5mu

| Seurat Component | h5mu Location | Notes |
|-----------------|--------------|-------|
| Each assay | `/mod/{modality}/` | One AnnData per modality |
| Assay counts | `/mod/{modality}/X` | Expression matrix |
| Assay metadata | `/mod/{modality}/var` | Feature metadata |
| Cell metadata | `/obs` | Global cell annotations |
| Reductions | Preserved via MuDataSeurat | |
| Graphs | Preserved via MuDataSeurat | |

### LoadH5MU: h5mu to Seurat

| h5mu Location | Seurat Component | Notes |
|--------------|-----------------|-------|
| `/mod/{modality}/` | Assay (renamed) | Modality -> assay name mapping |
| `/mod/{modality}/X` | Assay data | Expression matrix |
| `/obs` | `meta.data` | Global cell metadata merged |
| `/mod/{modality}/obsm/spatial` | Spatial coordinates | If present |

## Session Info

```{r sessionInfo}
sessionInfo()
```

```{r cleanup-final, echo=FALSE, results='hide'}
outs <- c("cbmc_multimodal.h5mu", "cbmc_roundtrip.h5mu")
existing_files <- outs[file.exists(outs)]
if (length(existing_files) > 0) file.remove(existing_files)
```
